Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-01-05 SIR_manuscript.tex
working on intro
updated intro

The observed dynamics of the COVID-19 epidemic is dirven by the underlying processes including both epidemiological processes (infection, recovery, etc.) and testing processes. In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the epidemic dynamics, and providing insights for the epidemic management/control.

Spereating the epidemiological processes from testing processes in a mechanistic modeling framework provides a tool to investigate the effect of the testing on the epidemic dynamics. Intutively, this effect is through isolation and quarantine practices. The informative/useful part is the sensitivity of the epidemic dynamics to testing, isolation and quarantine. Specifically, during COVID-19 epidemic, the isolation has applied to people who have been tested. Thus, the epidemic dynamics will depend on testing processes. The latter determines how tests are distributed across epidemiological compartments, or for a given testing intensity how many people in each compartment get tested. Specifically, in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be weighted equally across all compartments. However, in the case where tests are targeted certain clusters of the population (as might happen, for example, for diagnosis and treatment, screening people for access to things like flights, surgery, long-term-care facilities, etc., and for mitigation and control of the disease spread) the tests will be weighted differently, where the greater the testing weight for a compartment, the more likely people in that compartment get tested. \todo{edit the last sentence, it is very long!}

@@

It is often the purpose of testing, apart from the testing infrastructure, that dictates/defines \ali{[better word?]} the testing strategy. There are at least four different, and sometimes conflicting, purposes of testing: (i) diagnosis and treatment, (ii) screening people for access to things like flights, surgery, LTCs etc, (iii) surveillance, and (iv) mitigation and control of disease spread.

It is notable that, by random testing we mainly refer to the PCR test given that there is almost negligible virological testing for surveillance purposes. Furthermore, all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias. Specifically, when the purpose is "(i) diagnosis and treatment" or "(iv) mitigation and control of disease spread", the bias is towards infectious individuals more strongly for (i) than (iv). When people are tested for "(ii) screening", the bias is towards the group-specific characteristics [I am making this up for now!]. For example, people who are getting on flights may/will be more mobile than people awaiting for a surgery and are probably going to stay in long-term care homes and consequently more isolated to begin with.

>>>>

The observed dynamics of the COVID-19 epidemic is dirven by the underlying processes including both epidemiological processes (infection, recovery, etc.) and testing processes (who gets tested and intensity). In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the epidemic dynamics, and providing insights for the population-level epidemic control.

During COVID-19 epidemic, since isolation has applied to people including, but not limited to, those who have been tested, the epidemic dynamics will depend on who gets tested. That is, how tests are distributed across epidemiological compartments for a given testing intensity. Specifically, in the case where tests are randomly assigned within the population, the tests will be weighted equally across all compartments. Random testing is the simplest form of testing regime and might happen, for example, when disease surveillance is the purpose of testing. However, in the case where targeted testing is administered within the population, the testing will be more weighted towards the compartments with people who have higher probability of getting tested. Targeted testing might happen, for example, for different purposes including diagnosis and treatment, screening people for access to things like flights, surgery, long-term-care facilities, and for mitigation and control of the disease spread. Specifically, when the purpose is "diagnosis and treatment" or "mitigation and control of disease spread", the bias is skewed towards infectious individuals with even more weights for the diagnosis and treatment than the mitigation and control. Also, when people are tested for "screening", the bias is towards the group-specific characteristics \ali{better wording needed}. For example, people who are getting on flights may or will be more mobile than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently more isolated to begin with.

Furthermore, test-specific sensitivities, intensity and sample-to-answer reporting time are some components of the testing processes that influence the epidemic dynamic. We model these components and explore the trade-offs between them in achieving the epidemic control objective.

R0

Frequent proactive testing of asymptomatic individuals offers an additional way to reduce transmission, and is among the most powerful and least disruptive of non-pharmaceutical interventions such as face masks, social distancing.

@@

\ali{discussion maybe: all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias.}
